---
title: "Synthetic Parameter Recovery"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

for (pkg in c("devtools", "ggplot2", "posterior")) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop(
      "This notebook requires the suggested package `",
      pkg,
      "`. Install suggested dependencies before rendering."
    )
  }
}

ci_label <- function(alpha) {
  stopifnot(is.numeric(alpha), length(alpha) == 1L, !is.na(alpha))
  formatC(100.0 * alpha, width = 2L, format = "d", flag = "0")
}

build_detection_recovery_table <- function(fit, truth, alpha = 0.05) {
  alpha_label <- ci_label(alpha)
  estimates <- calculate_parameter_estimates(
    fit,
    mean_estimate = FALSE,
    alpha = alpha,
    include_description = FALSE
  )
  truth_values <- c(
    active_detection = truth$active_detection,
    passive_asymptomatic_detection = truth$passive_asymptomatic_detection,
    passive_symptomatic_detection = truth$passive_symptomatic_detection
  )
  labels <- c(
    active_detection = "Active detection",
    passive_asymptomatic_detection = "Passive asymptomatic detection",
    passive_symptomatic_detection = "Passive symptomatic detection"
  )
  output <- data.frame(
    group = "Detection",
    label = unname(labels[estimates$parameter]),
    parameter = estimates$parameter,
    truth = unname(truth_values[estimates$parameter]),
    posterior_median = estimates$median_estimate,
    lower_95 = estimates[[paste0("lower_", alpha_label)]],
    upper_95 = estimates[[paste0("upper_", alpha_label)]],
    stringsAsFactors = FALSE
  )
  output$truth_in_interval <- with(
    output,
    lower_95 <= truth & truth <= upper_95
  )
  output
}

build_severity_recovery_table <- function(fit, truth, alpha = 0.05) {
  alpha_label <- ci_label(alpha)
  estimates <- calculate_fatality_ratio(
    fit,
    mean_estimate = FALSE,
    alpha = alpha,
    naive_estimate = FALSE
  )
  output <- data.frame(
    group = "Severity",
    label = c("IFR", "SIR"),
    parameter = c("ifr", "sir"),
    truth = c(truth$ifr, truth$sir),
    posterior_median = c(
      estimates$ifr_median_estimate,
      estimates$sir_median_estimate
    ),
    lower_95 = c(
      estimates[[paste0("ifr_lower_", alpha_label)]],
      estimates[[paste0("sir_lower_", alpha_label)]]
    ),
    upper_95 = c(
      estimates[[paste0("ifr_upper_", alpha_label)]],
      estimates[[paste0("sir_upper_", alpha_label)]]
    ),
    stringsAsFactors = FALSE
  )
  output$truth_in_interval <- with(
    output,
    lower_95 <= truth & truth <= upper_95
  )
  output
}

build_recovery_table <- function(fit, truth, alpha = 0.05) {
  output <- rbind(
    build_detection_recovery_table(fit, truth, alpha = alpha),
    build_severity_recovery_table(fit, truth, alpha = alpha)
  )
  output$truth_in_interval_label <- ifelse(
    output$truth_in_interval,
    "Captured",
    "Missed"
  )
  output$truth_in_interval <- NULL
  output
}

plot_recovery_intervals <- function(recovery_table) {
  plot_data <- recovery_table
  plot_data$label <- factor(
    plot_data$label,
    levels = rev(plot_data$label)
  )
  ggplot2::ggplot(
    plot_data,
    ggplot2::aes(y = label)
  ) +
    ggplot2::geom_segment(
      ggplot2::aes(
        x = lower_95,
        xend = upper_95,
        yend = label,
        color = truth_in_interval_label
      ),
      linewidth = 1.1
    ) +
    ggplot2::geom_point(
      ggplot2::aes(
        x = posterior_median,
        color = truth_in_interval_label
      ),
      size = 2.5
    ) +
    ggplot2::geom_point(
      ggplot2::aes(x = truth),
      shape = 4,
      stroke = 1.1,
      size = 3,
      color = "black"
    ) +
    ggplot2::scale_color_manual(
      values = c(Captured = "#216869", Missed = "#B42318")
    ) +
    ggplot2::labs(
      x = "Probability",
      y = NULL,
      color = "95% CI captures truth",
      title = "Synthetic Truth vs 95% Credible Intervals"
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      legend.position = "bottom"
  )
}

notebook_dir <- dirname(normalizePath(knitr::current_input(dir = TRUE)))
project_root <- normalizePath(
  file.path(notebook_dir, "..", ".."),
  winslash = "/",
  mustWork = TRUE
)
devtools::load_all(project_root, quiet = TRUE, export_all = FALSE)
alpha <- 0.05
```

## Overview

This notebook is a heavier-weight companion to the package vignettes. Its goal
is to act like an integration check: generate a synthetic line list with known
parameters, fit the no-strata model with convergence-oriented Stan settings,
then compare the true generating values against the posterior 95% credible
intervals.

## Synthetic truth

The data-generating scenario is intentionally simple for this first notebook: a
single unstratified population observed weekly through both active and passive
surveillance.

```{r synthetic-truth}
analysis_seed <- 314L
truth <- list(
  population = 75000L,
  ifr = 0.06,
  sir = 0.45,
  active_detection = 0.18,
  passive_asymptomatic_detection = 0.08,
  passive_symptomatic_detection = 0.90
)
times <- seq(as.Date("2024-01-01"), by = "7 days", length.out = 12L)
force_of_infection <- matrix(
  data = 0.004,
  nrow = length(times),
  ncol = 1L
)
fit_args <- list(
  chains = 2L,
  iter = 1000L,
  seed = analysis_seed,
  cores = 1L,
  open_progress = interactive(),
  refresh = if (interactive()) 100L else 0L
)

truth_table <- data.frame(
  parameter = c(
    "population",
    "ifr",
    "sir",
    "active_detection",
    "passive_asymptomatic_detection",
    "passive_symptomatic_detection"
  ),
  value = unlist(truth, use.names = FALSE)
)
knitr::kable(truth_table, digits = 3)
```

## Simulate data

`create_sample_linelist()` accepts a single-row `population`/`sir`/`ifr` table
for this no-strata case.

```{r simulate-data}
linelist <- create_sample_linelist(
  strata = data.frame(
    population = truth$population,
    sir = truth$sir,
    ifr = truth$ifr
  ),
  times = times,
  active_detection = truth$active_detection,
  passive_asymptomatic_detection = truth$passive_asymptomatic_detection,
  passive_symptomatic_detection = truth$passive_symptomatic_detection,
  force_of_infection = force_of_infection,
  seed = analysis_seed
)

case_counts <- as.data.frame.matrix(
  stats::xtabs(~ detection + outcome, data = linelist)
)[, c("Asymptomatic", "Symptomatic", "Death")]

knitr::kable(case_counts)
```

## Build and fit the model

The priors are intentionally weak here because the goal is parameter recovery
from the synthetic data rather than prior sensitivity analysis.

```{r fit-model}
model <- SeverityEstimateModel(
  line_list = linelist,
  population = truth$population
) |>
  set_active_prior(alpha = 1.0, beta = 1.0) |>
  set_passive_asymptomatic_prior(alpha = 1.0, beta = 1.0) |>
  set_passive_symptomatic_prior(alpha = 1.0, beta = 1.0) |>
  set_timesteps("time") |>
  set_detection(
    "detection",
    map = c("Active" = "active", "Passive" = "passive")
  ) |>
  set_outcome(
    "outcome",
    map = c(
      "Asymptomatic" = "asymptomatic",
      "Symptomatic" = "symptomatic",
      "Death" = "severe"
    )
  )

estimate <- do.call(fit, c(list(model = model), fit_args))
```

## Sampler diagnostics

These are computed with `posterior::summarise_draws()` using
`posterior::default_convergence_measures()`.

```{r diagnostics}
draws <- posterior::as_draws_array(as.array(estimate@model_fit))
diagnostics <- posterior::summarise_draws(
  draws,
  posterior::default_convergence_measures()
)
knitr::kable(diagnostics, digits = 3)
```

## Parameter recovery

The table below is the core integration-style check: for each generating
parameter, compare the synthetic truth to the posterior median and 95% credible
interval.

```{r recovery-table}
recovery_table <- build_recovery_table(estimate, truth, alpha = alpha)

knitr::kable(recovery_table, digits = 3)
```

```{r recovery-plot, fig.height = 4.5, fig.width = 8}
plot_recovery_intervals(recovery_table)
```

If this notebook is being used as a manual integration check, the main things
to look for are:

1. No serious sampler pathologies in the diagnostics table.
2. Most or all parameters falling inside their 95% credible intervals.
3. Posterior medians landing reasonably close to the generating values.
