---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`SeverityEstimate` fits a Bayesian model to estimate the infection fatality
ratio (IFR) and symptomatic infection rate (SIR) from case line list data,
accounting for the fact that active and passive surveillance systems detect
cases at different rates and with different symptom profiles. The model is based
on [Lessler et al. (2016)](https://doi.org/10.1093/aje/kwv452).

The lightest workflow is to format your data with `time`, `detection`, and
`outcome` columns, plus any strata columns you want to model, then call
`default_model()` followed by `fit()`. For more customized models, you can
still build a `SeverityEstimateModel` manually and configure it with the
individual `set_*()` helpers.

```{r setup}
options(mc.cores = 1L)
library(SeverityEstimate)
```

## Sample data

We start by generating a synthetic line list with `create_sample_linelist()`.
This requires a `strata` table describing each subgroup's true underlying
parameters - the infection fatality ratio (`ifr`), the symptomatic infection
rate (`sir`), and the population size — along with a sequence of observation
times and detection probabilities for active and passive surveillance.

Here we define three age groups observed weekly over a 13-week period, with a
constant force of infection of 0.005 across all strata and time points.

```{r sample data}
strata <- do.call(rbind, lapply(list(
  list(age = "youth",  ifr = 0.25, sir = 0.50, population = 10000L),
  list(age = "adult",  ifr = 0.30, sir = 0.65, population = 20000L),
  list(age = "senior", ifr = 0.35, sir = 0.80, population = 5000L)
), as.data.frame))
times <- seq(as.Date("2024-01-01"), as.Date("2024-3-31"), "+7 days")
force_of_infection <- matrix(
  data = 0.005, nrow = length(times), ncol = nrow(strata)
)
linelist <- create_sample_linelist(
  strata,
  times,
  0.15,  # active detection probability
  0.05,  # passive detection probability for asymptomatic cases
  0.95,  # passive detection probability for symptomatic cases
  force_of_infection = force_of_infection,
  seed = 123L
)
head(linelist, n = 20L)
```

The resulting line list has one row per observed case, with columns for the
patient identifier, time step, age stratum, detection type (`Active` or
`Passive`), and outcome (`Asymptomatic`, `Symptomatic`, or `Death`).

```{r sample data summary}
summary(linelist)
```

## Building and fitting the model

`default_model()` is the shortest path from formatted data to a fitted model.
It expects the line list to contain `time`, `detection`, and `outcome`, and it
treats every other line-list column as a strata column with
`degrees_of_freedom = 0L`. Because the sample line list also contains a patient
identifier, we first drop that column before constructing the model.

`default_model()` also sets the weakly informative detection priors used
throughout this package:

- Active detection: `Beta(1, 1)`
- Passive asymptomatic detection: `Beta(1, 3)`
- Passive symptomatic detection: `Beta(3, 1)`

Once constructed, `fit()` samples from the package's precompiled Stan model,
passing any additional arguments (e.g. `chains`, `iter`) through to
`rstan::sampling()`. The settings below are intentionally small to keep this
vignette reasonably quick to render. They are not sufficient to assess
convergence or support serious inference. In practice you'd want to use more
chains and iterations and review the usual Stan diagnostics.

```{r fit model}
linelist_model_data <- linelist[, c("time", "age", "detection", "outcome")]
population <- strata[, c("age", "population")]

model <- default_model(linelist_model_data, population)
model

estimate <- fit(
  model,
  chains = 1L,
  iter = 250L,
  seed = 123L,
  refresh = 0
)
summary(estimate)
```

If you need more control, you can still start from `SeverityEstimateModel()`
and use `set_strata()`, `set_timesteps()`, `set_detection()`, `set_outcome()`,
and the prior setters directly. That is useful, for example, when you want a
smoothed ordered strata effect with `degrees_of_freedom > 0L` rather than the
default unsmoothed categorical effect.

## Extracting results

### Detection rate parameters

`calculate_parameter_estimates()` returns a summary of the three detection rate
parameters fitted by the model: the active detection rate, the passive
asymptomatic detection rate, and the passive symptomatic detection rate. The
`alpha` argument controls which credible interval bounds are returned.

```{r underlying parameters}
calculate_parameter_estimates(estimate, alpha = 0.05)
```

### Fatality ratios

`calculate_fatality_ratio()` returns IFR and SIR estimates broken down by
stratum. Setting `naive_estimate = TRUE` also includes a naive observed ratio
computed directly from the raw counts, which provides a useful point of
comparison against the model-adjusted estimates.

```{r severity estimates}
calculate_fatality_ratio(
  estimate, median_estimate = FALSE, naive_estimate = TRUE
)
```
