---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`SeverityEstimate` fits a Bayesian model to estimate the infection fatality
ratio (IFR) and symptomatic infection rate (SIR) from case line list data,
accounting for the fact that active and passive surveillance systems detect
cases at different rates and with different symptom profiles. The model is based
on [Lessler et al. (2016)](https://doi.org/10.1093/aje/kwv452).

The package uses a pipeline API similar in spirit to `tidymodels`: you first
build and configure a `SeverityEstimateModel` object, then call `fit()` to run
the Stan sampler and obtain a `SeverityEstimateFit` from which estimates are
extracted.

```{r setup}
options(mc.cores = 1L)
library(SeverityEstimate)
```

## Sample data

We start by generating a synthetic line list with `create_sample_linelist()`.
This requires a `strata` table describing each subgroup's true underlying
parameters - the infection fatality ratio (`ifr`), the symptomatic infection
rate (`sir`), and the population size — along with a sequence of observation
times and detection probabilities for active and passive surveillance.

Here we define three age groups observed weekly over a 13-week period, with a
constant force of infection of 0.005 across all strata and time points.

```{r sample data}
strata <- do.call(rbind, lapply(list(
  list(age = "youth",  ifr = 0.25, sir = 0.50, population = 10000L),
  list(age = "adult",  ifr = 0.30, sir = 0.65, population = 20000L),
  list(age = "senior", ifr = 0.35, sir = 0.80, population = 5000L)
), as.data.frame))
times <- seq(as.Date("2024-01-01"), as.Date("2024-3-31"), "+7 days")
force_of_infection <- matrix(
  data = 0.005, nrow = length(times), ncol = nrow(strata)
)
linelist <- create_sample_linelist(
  strata,
  times,
  0.15,  # active detection probability
  0.05,  # passive detection probability for asymptomatic cases
  0.95,  # passive detection probability for symptomatic cases
  force_of_infection = force_of_infection,
  seed = 123L
)
head(linelist, n = 20L)
```

The resulting line list has one row per observed case, with columns for the
patient identifier, time step, age stratum, detection type (`Active` or
`Passive`), and outcome (`Asymptomatic`, `Symptomatic`, or `Death`).

```{r sample data summary}
summary(linelist)
```

## Building and fitting the model

The modelling workflow has two stages. First, a `SeverityEstimateModel` is
constructed by specifying the line list, the population table, and the model
configuration. Each `set_*` call adds one piece of configuration and returns
the model so calls can be chained.

- `set_active_prior/passive_asymptomatic_prior/passive_symptomatic_prior`:
  Set Beta distribution priors for each surveillance detection rate. Here we use
  a weakly informative prior for active detection and priors that reflect the
  expectation that passive surveillance is more likely to capture symptomatic
  than asymptomatic cases.
- `set_strata`: Declares the column used for stratification and its ordered
  levels. Levels should be provided explicitly to ensure a consistent ordering
  in the output.
- `set_timesteps`: Declares the column that identifies the time period of each
  observation.
- `set_detection`: Maps the values in the detection column to the canonical
  types `active` and `passive` expected by the model.
- `set_outcome`: Maps the values in the outcome column to the canonical types
  `asymptomatic`, `symptomatic`, and `severe` expected by the model.

Once configured, `fit()` compiles the Stan model and runs the sampler, passing
any additional arguments (e.g. `chains`, `iter`) through to `rstan::stan()`.
The settings below are intentionally small to keep this vignette reasonably
quick to render. They are not sufficient to assess convergence or support
serious inference. In practice you'd want to use more chains and iterations
and review the usual Stan diagnostics.

```{r fit model}
population <- strata[, c("age", "population")]

model <- SeverityEstimateModel(linelist, population) |>
  set_active_prior(alpha = 1.0, beta = 1.0) |>
  set_passive_asymptomatic_prior(alpha = 1.0, beta = 3.0) |>
  set_passive_symptomatic_prior(alpha = 3.0, beta = 1.0) |>
  set_strata(
    "age",
    levels = strata[, "age"],
    degrees_of_freedom = 1L
  ) |>
  set_timesteps("time") |>
  set_detection(
    "detection",
    map = c("Active" = "active", "Passive" = "passive")
  ) |>
  set_outcome(
    "outcome",
    map = c(
      "Asymptomatic" = "asymptomatic",
      "Symptomatic" = "symptomatic",
      "Death" = "severe"
    )
  )

estimate <- fit(
  model,
  chains = 1L,
  iter = 500L,
  seed = 123L,
  refresh = 0
)
```

## Extracting results

### Detection rate parameters

`calculate_parameter_estimates()` returns a summary of the three detection rate
parameters fitted by the model: the active detection rate, the passive
asymptomatic detection rate, and the passive symptomatic detection rate. The
`alpha` argument controls which credible interval bounds are returned.

```{r underlying parameters}
calculate_parameter_estimates(estimate, alpha = 0.05)
```

### Fatality ratios

`calculate_fatality_ratio()` returns IFR and SIR estimates broken down by
stratum. Setting `naive_estimate = TRUE` also includes a naive observed ratio
computed directly from the raw counts, which provides a useful point of
comparison against the model-adjusted estimates.

```{r severity estimates}
ratios <- calculate_fatality_ratio(
  estimate, median_estimate = FALSE, naive_estimate = TRUE
)
ratios[match(c("youth", "adult", "senior"), ratios$age), ]
```
